Trajectories

### 2022-12-09 index.qmd
Update model formula and add DAG

---
title: "Trajectories"
subtitle: "Lab notes"
date: "Updated: `r format(Sys.Date(), '%Y/%m/%d')`"
authors:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    corresponding: true
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias fargas 25-27
        city: Barcelona
        state: Spain
        postal-code: 08005
  - name: Daniela S. Ávila-Varela
    orcid: 0000-0002-3518-8117
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations:
      - name: Universidad Atónoma de Madrid
        department: Departamento de Psicología
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
abstract: "Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those in the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months."
format:
    html:
        toc: true
        link-external-newwindow: true
        number-sections: true
        number-offset: 0
        code-fold: true
        code-overflow: scroll
        code-line-numbers: true
        code-copy: hover
        fig-dpi: 1000
        reference-location: margin
        execute-dir: "C:/Users/U155880/Documents/trajectories/"
    pdf:
        toc: true
        number-sections: true
        colorlinks: true
        fig-dpi: 1000
bibliography: "references.bib"
csl: "manuscript/apa6.csl"
---

@@

Data and materials are available at [OSF](https://osf.io/hy984/), and code is available at [GitHub](https://github.com/gongcastro/trajectories).

@@

```{r}
#| label: fig-participants-lp
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Participant degree of exposure (DoE) to Catalan and Spanish"
participants %>% 
    pivot_longer(
        c(doe_catalan, doe_spanish),
        names_to = "language", values_to = "doe", 
        names_transform = ~str_to_sentence(str_remove(., "doe_"))
    ) %>% 
    mutate(
        doe = cut(doe, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
        id = paste0(id, " (", time, ")")
    ) %>% 
    count(language, doe) %>% 
    ggplot(aes(doe, y = n, label = n, fill = language)) +
    geom_col(colour = "white", size = 1) +
    geom_text(position = position_nudge(y = 5)) +
    facet_wrap(~language, ncol = 1) + 
    labs(
        x = "Degree of exposure to Catalan",
        y = "Number of participants",
        colour = "Language",
        fill = "Language"
    ) +
    scale_fill_manual(values = clrs[c(1, 3, 5)]) +
    theme(
        legend.position = "top",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()
    )
```

@@

Word lexical frequency ($Frequency$): lexical frequency of the word in its corresponding language, expressed as Zipf scores [@van2014subtlex; @zipf1949human]. This variable ranges from 0 to 7, and follows and approximates a normal distribution, with most values in a corpus ranging from 3 to 5 points. Lexical frequencies were extracted from the English corpora of the [CHILDES](https://childes.talkbank.org/) database using the [{childesr}](https://langcog.github.io/childes-db-website/) R package. Available words in the corpora were then mapped to their Catalan and Spanish translations. Responses to words with missing lexical frequency score were excluded from analyses of the total number of items).

```{r}
#| label: fig-items-freq
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of lexical frequency across items"
items %>% 
    ggplot() +
    aes(
        x = language,
        y = freq_zipf,
        fill = language,
        colour = language
    ) +
    stat_slab(slab_colour = "white", slab_size = 1, alpha = 0.5, scale = 0.5) +
    stat_dots(side = "left", scale = 0.5, position = position_nudge(x = -0.05)) + 
    stat_pointinterval(
        position = position_dodge(width = 0.1),
        point_size = 2.5,
        colour = "black"
    ) +
    labs(
        x = "Frequency (Zipf)",
        y = "Probability density",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    theme(
        legend.position = "none",
        legend.title = element_blank()
    )
```

@@

```{r}
#| label: fig-items-nphon
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of number of phonemes across items"
items %>%
    count(language, n_phon) %>% 
    ggplot() +
    aes(
        x = n_phon,
        y = n,
        fill = language,
        colour = language,
        shape = language
    ) +
    facet_wrap(~language, ncol = 1) +
    geom_col() +
    geom_text(aes(label = n), position = position_nudge(y = 5), colour = "black") +
    labs(
        x = "# Phonemes",
        y = "Number of words",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_x_continuous(breaks = 1:15) +
    theme(
        legend.position = "none",
        axis.ticks.x = element_line(),
        legend.title = element_blank()
    )
```

@@

```{r}
#| label: fig-items-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 9
#| fig-cap: "Distribution of Levenshtein similarities across items"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    mutate(lv = cut(lv, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>% 
    count(lv) %>% 
    mutate(prop = n/sum(n)) %>% 
    ggplot(aes(x = lv, n)) +
    geom_col(fill = clrs[1], colour = "white") +
    geom_text(aes(label = paste0(n, " (", percent(prop, accuracy = 0.1), ")")),
              position = position_nudge(y = 3), size = 3.5) +
    labs(x = "Levenshtein distance", y = "Probability density") +
    theme(
        panel.grid.major.x = element_blank(),
        legend.position = "top"
    )
```

@@

```{r}
#| label: fig-items-frequency-phonemes
#| echo: true
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Association betwen number of lexical frequency and number of phonemes"
items %>% 
    drop_na(freq_zipf) %>% 
    count(n_phon, freq_zipf, language) %>% 
    ggplot(aes(n_phon, freq_zipf, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 2, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        x = "# Phonemes", 
        y = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

@@

```{r}
#| label: fig-items-frequency-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen lexical frequency and phonological similarity"
items %>% 
    left_join(select(bvqdev::pool, item, language)) %>% 
    drop_na(freq_zipf) %>% 
    count(freq_zipf, lv, language) %>% 
    ggplot(aes(freq_zipf, lv, colour = language, fill = language, size = n)) +
    facet_wrap(~language) +
    geom_point(size = 2, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        y = "Levenshtein similarity", 
        x = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid.major.y = element_blank()
    )
```

@@

```{r}
#| label: fig-items-nphon-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen number of phonemes and phonological similarity"
items %>% 
    left_join(select(bvqdev::pool, item, language)) %>% 
    drop_na(freq_zipf) %>% 
    count(n_phon, lv, language) %>% 
    ggplot(aes(n_phon, lv, colour = language, fill = language, size = n)) +
    facet_wrap(~language) +
    geom_point(shape = 1, stroke = 1) +
    geom_smooth(method = "lm", formula = "y ~ x", size = 1) +
    labs(
        x = "# Phonemes", 
        y = "Levenshtein similarity", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

@@

We created a predictor ($Exposure$) to account for the exposure rate of each child to each of the words their parents responded to, weighted by the child's exposure to the language the word belongs to. the exposure of the $i$-th child to the $j$-th word measure is the product of the lexical frequency of the word (Zipf score) and the child’s degree of exposure to the corresponding language (a proportion)

$$
\textbf{Exposure}_{ij} = \textbf{Frequency}_i \times \textbf{DoE}_{j}
$$

For instance, for a child who is reportedly exposure to Catalan 80% of the time, and to Spanish 20% of the time, the expected exposure to the word *cavall* (horse, in Catalan, with a lexical frequency of `r round(items$freq_zipf[items$item=="cat_cavall"], 2)`) would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.20, 2)`:

$$
\begin{aligned}
\textbf{Exposure}_{\textbf{cavall, Catalan}} &= \textbf{Frequency}_{cavall} \times \textbf{DoE}_{Catalan} = 6.14 \times 0.8 = 4.91 \\
\textbf{Exposure}_{\textbf{caballo, Spanish}} &= \textbf{Frequency}_{caballo} \times \textbf{DoE}_{Spanish} = 6.14 \times 0.2 = 1.23
\end{aligned}
$$

@@

@@

We extracted four variables of interest for each observation (describe in previous sections):

* Word lexical frequency ($Frequency$)
* Word length ($Phonemes$)
* Phonological similarity ($Levenshtein$)
* Participant age ($Age$)
* Word exposure index ($Exposure$)

We fitted a series of nested Bayesian ordinal multilevel models to sequentially test the effects of the independent variables (number of phonemes, exposure, age, and Levenshtein) over the three levels of word learning (no, production, comprehension). In all these models participants’ and items’ individual differences were controlled by estimating different parameters for each one (different intercepts and/or slopes). This is similar to the control conducted by ANCOVAs. Thus, independent variable effects account for collective differences in word learning (unexplained by participants’ or items’ individual differences).

We initially fitted a base model (`model_fit_0`) that only included the predictors $Age$ (`age_std`), $Phonemes$ (`n_phon_std`), along with random intercepts and slopes of all predictors by $Participant$ (`id`) and $Item$ (`te`), and all correlation parameters. We then expanded this model by adding the main effect of $Exposure$ (`exposure`) (`model_fit_1`), and then $Levenshtein$ (`lv_std`) (`model_fit_2`) and their corresponding random slopes by $Participant$. Then, we added the $Exposure \times Levenshtein$ (`exposure_std:lv_std`) two-way interaction (`model_fit_3`) and its random slope by $Participant$. Finally, we added the $Age \times Exposure \times Levenshtein$ (`age_std:exposure_std:lv_std`) three-way interaction with its random slope by $Participant$, along with a random slope for the $Age \times Exposure$ (`age_std:exposure_std`) two-way interaction.

We compared the predictive performance of the models using Bayesian leave-one-out cross-validation (*LOO*), a relative fit index, via the `loo` R package. LOO consists in computing the average likelihood of each observation after estimating the model's parameters leaving that same observation out of the data set. Due to computational constraints, we performed LOO based on a random subsample of 750 samples (out of `r format(dim(model_fit_4$fit)[1], big.mark = ",")`) of the posterior distribution of the fixed effects of the model.

### Model equation

::: column-body-outset
$$\begin{aligned}
g[\mathrm{Pr}(y_{ij} \le a_s)] = \tau_s - \nu_{ij}
\end{aligned}$$
:::

$s$ is the s-th response category, $a_1 < a_2 < a_3$ are ordered response categories, $\mathrm{Pr}(y_i \le a_3) = 1$ and $\tau_s$ are threshold parameters where $\tau_1 < \tau_2 < \tau_3$.

::: column-body-outset

$$\begin{aligned}
\nu_{ij} = & \beta_0 +\beta_1\mathrm{Age}_{i} + \beta_2\mathrm{Phon}_{j} + \beta_3\mathrm{Exposure}_{ij} + \beta_4 \mathrm{Levenshtein}_{j} + & [\mbox{Main effects}]\\
&\beta_5\mathrm{(Age_{i} \cdot \mathrm{Exposure}_{ij}}) + \beta_6\mathrm{(\mathrm{Age}_{i} \cdot \mathrm{Levenshtein}_{j}}) +  \beta_7\mathrm{(Exposure_{ij} \cdot \mathrm{Levenshtein}_{j}}) + & [\mbox{Two-way interactions}]\\
&\beta_8{(\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij} \cdot \mathrm{Levenshtein}_{j}}) + & [\mbox{Three-way interaction}]\\

&\theta_{0j}^{(\mathrm{TE})} + \theta_{1j}^{(\mathrm{TE})}\mathrm{Age}_{i} + \theta_{2j}^{(\mathrm{TE})} \mathrm{Phon}_{j} + \theta_{3j}^{(\mathrm{TE})} \mathrm{Exposure}_{ij} +  \theta_{4j}^{(\mathrm{TE})} (\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij}) + & [\mbox{Item-level effects}]\\

&\theta_{0j}^{(\mathrm{ID})} + \theta_{1j}^{(\mathrm{ID})}\mathrm{Age}_{i} +   \theta_{2j}^{(\mathrm{ID})} \mathrm{Phon}_{j} + \theta_{3j}^{(\mathrm{ID})} \mathrm{Exposure}_{ij} + \theta_{4j}^{(\mathrm{ID})} \mathrm{Levenshtein}_{j} + \theta_{5j}^{(\mathrm{ID})} (\mathrm{Age}_{i} \cdot \mathrm{Exposure}_{ij}) & [\mbox{Participant-level effects}]
\end{aligned}$$
:::

### Model prior

::: column-body-outset
$$\begin{aligned}\beta_{0} & \sim \mathrm{Normal}(-0.25, 0.1) & [\mbox{Intercept/response category threshold}] \\
\beta_{1} &\sim \mathrm{Normal}(1, 0.1) & [\mbox{Age population-level coefficient}]\\
\beta_{2-9} &\sim \mathrm{Normal}(0, 1) & [\mbox{Rest of population-level coefficients}] \\
\theta_{0-9} & \sim \mathrm{Normal}(1, 0.1) & [\mbox{Group-level coefficient variability}]\end{aligned}$$
:::

In addition, we specified a $LKJ$ prior--$\sim LKJ(2)$ for the correlations between group-level coefficients.

>>>>

---
title: "Trajectories"
subtitle: "Lab notes"
date: "Updated: `r format(Sys.Date(), '%Y/%m/%d')`"
authors:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    corresponding: true
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias fargas 25-27
        city: Barcelona
        state: Spain
        postal-code: 08005
  - name: Daniela S. Ávila-Varela
    orcid: 0000-0002-3518-8117
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations:
      - name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
  - name: Nuria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations:
      - name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
abstract: "Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those in the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months."
format:
    html:
        toc: true
        link-external-newwindow: true
        number-sections: true
        number-offset: 0
        code-fold: true
        code-overflow: scroll
        code-line-numbers: true
        code-copy: hover
        fig-dpi: 1000
        reference-location: margin
        execute-dir: "C:/Users/U155880/Documents/trajectories/"
    pdf:
        toc: true
        number-sections: true
        colorlinks: true
        fig-dpi: 1000
bibliography: "references.bib"
csl: "manuscript/apa6.csl"
---

@@

@@

```{r}
#| label: fig-participants-lp
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Participant degree of exposure (DoE) to Catalan and Spanish"
participants %>% 
    pivot_longer(
        c(doe_catalan, doe_spanish),
        names_to = "language", values_to = "doe", 
        names_transform = ~str_to_sentence(str_remove(., "doe_"))
    ) %>% 
    mutate(
        doe = cut(doe, breaks = seq(0, 1, 0.1), include.lowest = TRUE),
        id = paste0(id, " (", time, ")")
    ) %>% 
    count(language, doe) %>% 
    ggplot(aes(doe, y = n, label = n, fill = language)) +
    geom_col(colour = "white", size = 1) +
    geom_text(position = position_nudge(y = 5)) +
    facet_wrap(~language, ncol = 1) + 
    labs(
        x = "Degree of exposure to Catalan",
        y = "Number of participants",
        colour = "Language",
        fill = "Language"
    ) +
    scale_fill_manual(values = clrs[c(1, 4)]) +
    theme(
        legend.position = "top",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()
    )
```

@@

Word lexical frequency ($Frequency$): lexical frequency of the word in its corresponding language, expressed as Zipf scores [@van2014subtlex; @zipf1949human]. This variable ranges from 0 to 7, and follows and approximates a normal distribution, with most values in a corpus ranging from 3 to 5 points. Lexical frequencies were extracted from the English corpora of the [CHILDES](https://childes.talkbank.org/) database using the [{childesr}](https://langcog.github.io/childes-db-website/) R package. Available words in the corpora were then mapped to their Catalan and Spanish translations. Therefore, the Catalan and Spanish word-forms of the same translation equivalent had the same value for lexical frequency. Responses to words with missing lexical frequencies were excluded from analyses of the total number of items).

```{r}
#| label: fig-items-freq
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Distribution of lexical frequency across items. Lexical frequencies were extracted from the English CHILDES corpora, and mapped to their Catalan and Spanish translations. Therefore Catalan and Spanish word-forms that belong to the same translation equivalent had the same lexical frequency. The black dot and interval indicate the mean and standard deviation."

mean_sd <- function(x, ...) {
    m <- mean(x, ...)
    s <- sd(x, ...)
    data.frame(mean = m, ymin = m-s, ymax = m+s)
}

items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    ggplot(aes(x = freq_zipf)) +
    geom_histogram(fill = clrs[4], colour = "white", bins = 30)  +
    labs(x = "Lexical frequency (Zipf score)", y = "Number of words") +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        legend.title = element_blank()
    )
```

@@

```{r}
#| label: fig-items-nphon
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Distribution of number of phonemes across items"
items %>%
    count(language, n_phon) %>% 
    ggplot() +
    aes(
        x = n_phon,
        y = n,
        fill = language,
        colour = language,
        shape = language
    ) +
    facet_wrap(~language, ncol = 1) +
    geom_col(colour = "white") +
    geom_text(aes(label = n), position = position_nudge(y = 5), colour = "black") +
    labs(
        x = "# Phonemes",
        y = "Number of words",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_x_continuous(breaks = 1:15) +
    scale_fill_manual(values = clrs[c(1, 4)]) +
    theme(
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.x = element_line(),
        legend.title = element_blank()
    )
```

@@

```{r}
#| label: fig-items-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 9
#| fig-cap: "Distribution of Levenshtein similarities across items"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    mutate(lv = cut(lv, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>% 
    count(lv) %>% 
    mutate(prop = n/sum(n)) %>% 
    ggplot(aes(x = lv, n)) +
    geom_col(fill = clrs[4], colour = "white") +
    geom_text(aes(label = paste0(n, " (", percent(prop, accuracy = 0.1), ")")),
              position = position_nudge(y = 3), size = 3.5) +
    labs(x = "Levenshtein distance", y = "Number of words") +
    theme(
        panel.grid.major.x = element_blank(),
        legend.position = "top"
    )
```

@@

```{r}
#| label: fig-items-frequency-phonemes
#| echo: true
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Association betwen number of lexical frequency and number of phonemes"
items %>% 
    drop_na(freq_zipf) %>% 
    distinct(te, .keep_all = TRUE) %>% 
    count(n_phon, freq_zipf) %>% 
    ggplot(aes(n_phon, freq_zipf)) +
    geom_point(size = 2, alpha = 0.5, colour = clrs[4], shape = 1, stroke = 1,
               position = position_jitter(width = 0.1), seed = 888) +
    geom_smooth(method = "lm", formula = "y ~ x" ,
                colour = clrs[1], fill = clrs[1]) +
    labs(
        x = "# Phonemes", 
        y = "Lexical frequency (Zipf score)", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

@@

```{r}
#| label: fig-items-frequency-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association betwen lexical frequency and phonological similarity"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    drop_na(freq_zipf) %>% 
    count(freq_zipf, lv) %>% 
    ggplot(aes(freq_zipf, lv, size = n)) +
    geom_point(size = 2, shape = 1, stroke = 1, colour = clrs[4]) +
    geom_smooth(method = "lm", formula = "y ~ x", colour = clrs[1], fill = clrs[1]) +
    labs(
        y = "Levenshtein similarity", 
        x = "Lexical frequency (Zipf score)", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid.major.y = element_blank()
    )
```

@@

```{r}
#| label: fig-items-nphon-lv
#| echo: true
#| message: false
#| warning: false
#| fig-width: 7
#| fig-cap: "Association between number of phonemes and phonological similarity"
items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    drop_na(freq_zipf) %>% 
    count(n_phon, lv) %>% 
    ggplot(aes(n_phon, lv, size = n)) +
    geom_point(shape = 1, stroke = 1, colour = clrs[4]) +
    geom_smooth(method = "lm", formula = "y ~ x", size = 1,
                colour = clrs[1], fill = clrs[1]) +
    labs(
        x = "# Phonemes", 
        y = "Levenshtein similarity", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

@@

We created a predictor ($Exposure$) to account for the exposure rate of each child to each of the words their parents responded to, weighted by the child's exposure to the language the word belongs to. the exposure of the $i$-th child to the $j$-th word measure is the product of the lexical frequency of the word (Zipf score) and the child’s degree of exposure to the corresponding language (a proportion) (see @eq-exposure).

$$
\textbf{Exposure}_{ij} = \textbf{Frequency}_i \times \textbf{DoE}_{j}
$$ {#eq-exposure}

For instance, for a child who is reportedly exposure to Catalan 80% of the time, and to Spanish 20% of the time, the expected exposure to the word *cavall* (horse, in Catalan, with a lexical frequency of `r round(items$freq_zipf[items$item=="cat_cavall"], 2)`) would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq_zipf[items$item=="cat_cavall"]*0.20, 2)` (see @eq-exposure-example):

$$
\begin{aligned}
\textbf{Exposure}_{\textbf{cavall, Catalan}} &= \textbf{Frequency}_{cavall} \times \textbf{DoE}_{Catalan} = 6.14 \times 0.8 = 4.91 \\
\textbf{Exposure}_{\textbf{caballo, Spanish}} &= \textbf{Frequency}_{caballo} \times \textbf{DoE}_{Spanish} = 6.14 \times 0.2 = 1.23
\end{aligned}
$$ {#eq-exposure-example}

@@

## DAG

```{r tikz-settings, include=FALSE}
# tikz stuff
# Necessary for using dvisvgm on macOS
# See https://www.andrewheiss.com/blog/2021/08/27/tikz-knitr-html-svg-fun/
font_opts <- list(dvisvgm.opts = "--font-format=woff")
```

```{tikz dag, engine.opts=font_opts, echo=FALSE}
#| fig-cap: "Directed acyclic graph indicating the hypothesised causal relationship between the variables of interest. *i* and *j* indicate the variables that are affected by individual differences at the level of participants and translation equivalents, respectively. Unobserved variables are encircled, and observed variables are not. We hypothesised that the number of phonemes of the word, the age of the participant, the amount of exposure to the word, and the Levenshtein distance all affect participants' familiarity with the word (i.e., the probability of acquisition), and therefore the probability with which each response category is generated. We further hypothesised that age influences the effect of word exposure and Levenshtein distance on familiarity, and that word exposure influences the effect of Levenshtein distance on familiarity."
#| fig-align: center
#| fig-ext: svg
#| out-width: 100%
\usetikzlibrary{positioning}
\usetikzlibrary{shapes.geometric}
\begin{tikzpicture}
% nodes
\node (age) at (0,0) {$\text{Age}_{i, j}$};
\node (phonemes) at (3,4) {$\text{Phonemes}_{i, j}$};
\node (exposure) at (3,2) {$\text{Exposure}_{i, j}$};
\node (levenshtein) at (3,-2) {$\text{Levenshtein}_{i}$};
\node [ellipse, draw](familiarity) at (7, 0) {$Familiarity$};
\node (response) at (11, 0) {$\text{Response}$};
% arrows
\draw [->, shorten >= 4pt, shorten <= 4pt] (age) -- (familiarity);
\draw [->, shorten >= 4pt, shorten <= 4pt] (age) -- (exposure);
\draw [->, shorten >= 4pt, shorten <= 4pt] (age) -- (levenshtein);
\draw [->, shorten >= 4pt, shorten <= 4pt] (exposure) -- (familiarity);
\draw [->, shorten >= 4pt, shorten <= 4pt] (exposure) -- (levenshtein);
\draw [->, shorten >= 4pt, shorten <= 4pt] (phonemes) -- (familiarity);
\draw [->, shorten >= 4pt, shorten <= 4pt] (levenshtein) -- (familiarity);
\draw [->, shorten >= 4pt, shorten <= 4pt] (familiarity) -- (response);

\end{tikzpicture}
```

@@

We used multilevel ordinal regression to model the cumulative probability of a *No* response, a *Understands* response, or a *Understands and Says* response [@burkner2019ordinal] using the *logit* link function. The likelihood of the cumulative probability distribution of the responses is defined by @eq-likelihood (see @fig-likelihood for a graphical representation).

$$
\begin{aligned}
\textbf{Likelihood:} \\
y_{ij} &\sim \text{Cumulative}(p_{k})
\end{aligned}
$$ {#eq-likelihood}

where:

- $y$ is an observed response ($y \in \{\text{No, Understands, Understands and Says}\}$)
- $i$ is the participant index
- $j$ is the translation equivalent (TE) index
- $p_{k}$ is a probability ($p \in (0, 1)$) that indicates the threshold $k$ ($k \in (1, 2)$)) between two response categories in the latent distribution
$p_{k}$ is then estimated using a logit regression model as indicated in @eq-linear.

```{r}
#| label: fig-likelihood
#| fig-cap: "Graphical representation of the assumed latent distribution that generates observed responses in the questionnaire. Cumulative ordinal regression assumes that observed categorical responses (*No*, *Understands* and *Understands and Says* are generated from a common latent (i.e., unobserved) distribution. For threotical reasons, we refer to the dimension captured by such distribution as *Familiarity*. *Familiarity* describes a given participant's familiarity with a given word. When familiarity is low (below some threshold 1), the reulting response is *No*. When familiarity is higher than such threshold 1, but lower than some threshold 2, the resulting result is *Understands* (i.e., comprehension). When familiarity exceeds threshold 2, the resulting response is *Understands and Says* (i.e., comprehension and production). These thresholds are estimated in the cumulative regression model as intercepts 1, and 2. The rest of the coefficients in the regression model shift the response thresholds, making comprehension or production more likely or less likely. See @burkner2019ordinal for a detailed description of cumulative ordinal regression and its implementation."
#| fig-width: 7
#| fig-height: 3.5
dist_df = tribble(~mean, ~sd, 0,   1)

responses <- data.frame(
    ymin = c(-Inf, -1.5, 1.5),
    ymax = c(-1.5, 1.5, Inf),
    xmin = -Inf,
    xmax = Inf,
    .category = str_wrap(c("No", "Understands", "Understands and Says"), width = 12)
)

responses %>%
    ggplot(aes(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax, fill = .category)) +
    geom_rect(alpha = 0.5) +
    stat_dist_slab(
        data = dist_df, 
        aes(dist = "norm", arg1 = mean, arg2 = sd),
        position = "dodge", inherit.aes = FALSE,
        fill = NA, colour = "black"
    ) +
    geom_hline(yintercept = c(-1.5, 1.5), linetype = "dotted", size = 1) +
    geom_text(aes(label = .category, y = c(-3, 0, 3), x = 1),
              vjust = 1, hjust = 0.5) +
    annotate(
        geom = "text",
        label = c("Threshold~1~(p[1])", "Threshold~2~(p[2])"),
        y = c(-1.5, 2), parse = TRUE,
        x = 0.5, angle = 90, vjust = -0.5
    ) +
    labs(y = "Hypothetical response-generating latent distribution",
         x = "Response probability",
         fill = "Response type",
         colour = "Response type") +
    coord_flip() +
    scale_fill_manual(values = clrs[c(1, 4, 5)]) +
    scale_x_continuous(labels = percent) +
    scale_y_continuous(limits = c(-4, 4)) +
    theme(panel.grid = element_blank(),
          legend.position = "none")
```

To test our hypotheses, we included several predictors in the regression model as fixed effects: the main effects of participant age ($Age$), number of phonemes ($Phonemes$), word exposure index ($Exposure$), and of phonological similarity ($Levenshtein$), the two-way interactions $Age \times Exposure$, $Age \times Levenshtein$, and $Exposure \times Leveshtein$, and the three-way interaction $Age \times Exposure \times Leveshtein$. We also included crossed random effects for participants and translation equivalents to account for the repeated measures in our dataset--each participant provided responses to multiple translation equivalents, and each translation equivalent was responded to by multiple participants [@gelman2020regression]. For both grouping variables, we included random intercepts, random slopes, and correlation parameters for all predictors were repeated measures were observed in our dataset [@barr2013random, see @eq-linear].

$$
\begin{aligned}
\textbf{Linear model:} \\
logit(p_{k}) = \text{ln} \frac{p_{k}}{1-p_{k}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + \\
& (\beta_{1} + u_{1_{i}} + w_{1_{j}}) · \text{Age}_{i} + & \\
& (\beta_{2} + u_{2_{i}} + w_{2_{j}}) · \text{Phonemes}_{ij} + & \\
& (\beta_{3} + u_{3_{i}} + w_{3_{j}}) · \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) · \text{Levenshtein}_{ij} + & \\
& (\beta_{5} + u_{5_{i}} + w_{3_{j}}) · (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) · (\text{Age}_{i} \times \text{Levenshtein}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) · (\text{Exposure}_{ij} \times \text{Levenshtein}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) · (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Levenshtein}_{ij}) & \\
\end{aligned}
$$ {#eq-linear}

where:
